Transfer learning for echelle spectroscopy

### 2021-11-16 paper/paper1/ms.tex
Missing word in the tex
Add placeholders for figures and pseudoVoigt subsection

\documentclass[modern]{aastex631}
\bibliographystyle{aasjournal}
\turnoffedit
\usepackage[caption=false]{subfig}
\usepackage{booktabs}
\usepackage{censor}
\usepackage{mathtools}

\usepackage{outlines}
\usepackage{amsmath,bm}

@@

Figure XX shows how the number of detected lines $N_{\mathrm{lines}}$ scales with effective temperature and prominence threshold $P_{rom}$ for the \texttt{PHOENIX} grid, truncated to the bandwidths-plus-buffers for HPF and IGRINS.  The HPF-truncated spectra have $N_s=XX$ and the IGRINS-truncated spectra have $N_s=XX$.

@@

The model evaluated with its coarse initial values would have terrible performance: it would only vaguely resemble the synthetic spectral model, with up to $\pm 50\%$ undulations from the inexact assignment of widths, lineshapes, and amplitudes.  Instead, we tune the parameters of the model, starting from these coarse initial values.  This model has $N_{\mathrm{lines}}\times 3$ free parameters, where the center wavelength is held fixed and the amplitude, width, and lineshape are allowed to vary.  We minimize a scalar ``goodness-of-fit'' metric, \emph{aka} loss scalar $\mathcal{L}$, chosen as the mean squared error (MSE), which is proportional to $\chi^2$, the sum of the squares of the residual vector $\mathsf{R} \equiv \mathsf{S}-\mathsf{M}$ but has no notion of per-pixel noise, since the precomputed synthetic spectrum has no uncertainty:

@@

As seen in Figure XX, the number of lines can exceed $XX$, meaning the model has over $XX \times 3 = ZZ$ free parameters.  Fitting that large number of parameters is difficult with conventional optimizers, which struggle to converge after hundreds of parameters\authorcomment1{Citation to limits of optimizers}.

@@

In principle we could compute the Jacobians analytically---since the partial derivatives of the Voigt function are known in closed-form---circumventing the vendor lock-in and memory overhead of an autodiff implementation with PyTorch.  Futher, an autodiff-aware implementation of the \emph{celerit\'e} algorithm\authorcomment1{add DFM citation} does not yet exist in PyTorch, constricting our choices for simultaneous continuum fitting.  In practice, PyTorch also offers hardware acceleration (on GPU or TPU)\authorcomment1{citation needed}, mature support for a range of optimizers, and other perks that motivated its choice over analytic implementations or other machine learning frameworks.  The GPyTorch framework\authorcomment1{add citation} may offer a workaround to the absence of a \emph{celerit\'e} implementation.  Still, the performance of an analytic Jacobian implementation over the autodiff route presented here offers an interesting avenue of future research.

@@

Figure XX shows a representative Loss curve (MSE versus epochs) for cloning the model in Figure ZZ.  The NN epochs of training took X minutes on an RTX2070 GPU with PyTorch v1.11.x, CUDA v11.x, and Intel X CPU, with all tensors as FP64.  That benchmark works out to X seconds per epoch, or about Z seconds per minibatch.

@@

We compute the residual $\mathsf{R}(\bm{\lambda}_s)$ of native PHOENIX minus cloned model, illustrated in the bottom panel of Figure XX\authorcomment1{make this figure}.  We see an RMS of $\blackout{1}\%$ residuals at native resolution, with a long tail of residuals up to $\blackout{20}\%$.  The greatest residuals surround the deepest and broadest lines---such as Hydrogen and neutral alkali metal lines---for which the non-Voigt lineshapes stand out.  Figure ZZ\authorcomment1{make this figure} zooms in on the XX\authorcomment1{todo} line at $\censor{10200}\;$\AA.

@@

\subsection{Line data clustering and conspicuous flaws}
We see some narrow lines devolve into missing continuum opacity, the favored tradeoff when the continuum estimate's poor performance outweighs the the pain of a narrow-but-tolerably-small spike.  This problem can be seen in Figure XX where a line initialized at $\censor{10210}\;$\AA~ with an amplitude of ZZ and width of YY bloats to \censor{some other value}.  These devolved lines tend to appear poorly immitated line wings.  One can mitigate these devolved lines by adding more flexibility to the continuum or supporting more advanced line shapes.

>>>>

\documentclass[modern]{aastex631}
\bibliographystyle{aasjournal}
\turnoffedit
\usepackage[caption=false]{subfig}
\usepackage{booktabs}
\usepackage{censor}
\usepackage{mathtools}
\usepackage{outlines}
\usepackage{amsmath,bm}

@@

\begin{figure}[hbt!]
  \centering
  \includegraphics[width=0.5\textwidth]{example-image-a}
  \caption{Number of prominent spectral lines versus effective temperature for PHOENIX models truncated to HPF (\emph{left}) and IGRINS (\emph{right}) bandwidths, for different prominence thresholds of $0.05$ and $0.01$.}
  \label{fig_Nlines_vs_teff}
\end{figure}

Figure \ref{fig_Nlines_vs_teff} shows how the number of detected lines $N_{\mathrm{lines}}$ scales with effective temperature and prominence threshold $P_{rom}$ for the \texttt{PHOENIX} grid, truncated to the bandwidths-plus-buffers for HPF and IGRINS.  The HPF-truncated spectra have $N_s=XX$ and the IGRINS-truncated spectra have $N_s=XX$.

@@

\subsection{Voigt profile evaluation}
The Voigt profile $\mathsf{V}(\lambda, \sigma_j, \gamma_j)$ can be computed in exact closed form using the Fadeeva function\authorcomment1{citation}.  Evaluation of the Fadeeva function can be computationally costly, and so approximate forms may be desirable.  Here we adopt the pseudo-Voigt approximation described in \censor{some paper}\authorcomment1{look up reference to PseudoVoigt}.

Two sampled Voigt approaches could be feasible.  Direct convolution of a sampled Gaussian and Lorentzian profile may be too costly and inexact if pixels are coarsely sampled.  Analogously, the convolution can be achieved in the Fourier domain through multiplication of the Fast Fourier Transforms of the Gaussian and Lorentzian profiles, followed by the subsequent Inverse Fourier Transforms.  The main rationale for either of these two choices would be to anticipate and align with planned convolutions such as rotational broadening and instrumental kernels.  We do not currently implement these sampled convolutions in \emph{blas\'e} due to their prohibitive computational cost.

\subsection{Goodness of fit metric}
The model evaluated with its coarse initial values would have terrible performance: it would only vaguely resemble the synthetic spectral model, with up to $\pm 50\%$ undulations from the inexact assignment of widths, lineshapes, and amplitudes.  Instead, we tune the parameters of the model, starting from these coarse initial values.  This model has $N_{\mathrm{lines}}\times 3$ free parameters, where the center wavelength is held fixed and the amplitude, width, and lineshape are allowed to vary.  We minimize a scalar ``goodness-of-fit'' metric, \emph{aka} loss scalar $\mathcal{L}$, chosen as the mean squared error (MSE), which is proportional to $\chi^2$, the sum of the squares of the residual vector $\mathsf{R} \equiv \mathsf{S}-\mathsf{M}$ but has no notion of per-pixel noise, since the precomputed synthetic spectrum has no uncertainty:

@@

As seen in Figure \ref{fig_Nlines_vs_teff}, the number of lines can exceed $XX$, meaning the model has over $XX \times 3 = ZZ$ free parameters.  Fitting that large number of parameters is difficult with conventional optimizers, which struggle to converge after hundreds of parameters\authorcomment1{Citation to limits of optimizers}.

@@

In principle we could compute the Jacobians analytically---since the partial derivatives of the Voigt function are known in closed-form---circumventing the vendor lock-in and memory overhead of an autodiff implementation with PyTorch.  Further, an autodiff-aware implementation of the \emph{celerit\'e} algorithm\authorcomment1{add DFM citation} does not yet exist in PyTorch, constricting our choices for simultaneous continuum fitting.  In practice, PyTorch also offers hardware acceleration (on GPU or TPU)\authorcomment1{citation needed}, mature support for a range of optimizers, and other perks that motivated its choice over analytic implementations or other machine learning frameworks.  The GPyTorch framework\authorcomment1{add citation} may offer a workaround to the absence of a \emph{celerit\'e} implementation.  Still, the performance of an analytic Jacobian implementation over the autodiff route presented here offers an interesting avenue of future research.

@@

\begin{figure}[hbt!]
  \centering
  \includegraphics[width=0.6\textwidth]{example-image-golden}
  \caption{PHOENIX spectrum cloned with blas\'e.}
  \label{fig_cloned_spectrum_demo}
\end{figure}

Figure \ref{fig_cloned_spectrum_demo} shows a portion of a PHOENIX spectrum cloned with \emph{blas'e}.  The $NN$ epochs of training took X minutes on an RTX2070 GPU with PyTorch v1.11.x, CUDA v11.x, and Intel X CPU, with all tensors as FP64.  That benchmark works out to X seconds per epoch, or about Z seconds per minibatch.

@@

We compute the residual $\mathsf{R}(\bm{\lambda}_s)$ of native PHOENIX minus cloned model, illustrated in the bottom panel of Figure \ref{fig_cloned_spectrum_demo}\authorcomment1{make this figure!}.  We see an RMS of $\blackout{1}\%$ residuals at native resolution, with a long tail of residuals up to $\blackout{20}\%$.  The greatest residuals surround the deepest and broadest lines---such as Hydrogen and neutral alkali metal lines---for which the non-Voigt lineshapes stand out.  Figure \ref{fig_poor_cloning_performance}\authorcomment1{make this figure} zooms in on the \censor{XX} line at $\censor{10200}\;$\AA.

\begin{figure}[hbt!]
  \centering
  \includegraphics[width=0.3\textwidth]{example-grid-100x100pt}
  \caption{A \censor{Hydrogen} line with broad line wings exhibiting poor cloning performance.}
  \label{fig_poor_cloning_performance}
\end{figure}

@@

\subsection{Line data clustering and conspicuous flaws}
We see some narrow lines devolve into missing continuum opacity, the favored tradeoff when the continuum estimate's poor performance outweighs the the pain of a narrow-but-tolerably-small spike.  This problem can be seen in Figure \ref{fig_poor_cloning_performance} where a line initialized at $\censor{10210}\;$\AA~ with an amplitude of ZZ and width of YY bloats to \censor{some other value}.  These devolved lines tend to appear at poorly immitated broad-and-deep line wings.  One can mitigate these devolved lines by adding more flexibility to the continuum or supporting more advanced line shapes.
